---
title: "Minimum distance of market places"
author: "Irache Vergara"
date: "2025-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r setup1, echo=TRUE, message=FALSE, warning=FALSE}
# Clean the environment
rm = list(ls())
rm(list = ls())

# Library
library(readxl)
library(sf)
library(ggplot2)
library(maps)
library(dplyr)
library(readr)
library(rnaturalearth)

```

## Prepare data

```{r setup2}
# Download and read the market coordinates data from a GitHub raw file
url <- "https://raw.githubusercontent.com/Viktoriag27/Geospatial_Assignment_2/main/Data/Part%202/MktCoords.xlsx"
temp_file <- tempfile(fileext = ".xlsx")  
download.file(url, temp_file, mode = "wb") 
market_data <- read_excel(temp_file)
```

```{r setup3}
# Convert the market coordinates dataframe into a spatial sf object using longitude and latitude
market_sf <- st_as_sf(market_data, coords = c("longitude", "latitude"), crs = 4326)
print(market_sf)

world_map <- map_data("world")
```


```{r setup4}
# Filter for Africa’s 
africa_map <- subset(world_map, long > -25 & long < 60 & lat > -40 & lat < 40)

# Create plot
ggplot() +
  geom_polygon(data = africa_map, aes(x = long, y = lat, group = group), fill = "gray90", color = "white") +
  geom_sf(data = market_sf, color = "red", size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Market Locations") +  # Remove x and y labels
  theme(
    axis.title = element_blank(),  # Remove axis titles
    axis.text = element_blank(),   # Remove axis text
    axis.ticks = element_blank(),   # Remove axis ticks
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
```


```{r setup5}
# Download and read the price data from a GitHub raw file
price_file <- "https://raw.githubusercontent.com/Viktoriag27/Geospatial_Assignment_2/main/Data/Part%202/PriceMaster4GAMS.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
download.file(price_file, temp_file, mode = "wb")
price_data <- read_excel(temp_file)

head(price_data)
```


```{r setup6}
# Convert price columns to numeric format, ignoring the first 4 columns 
price_data <- price_data %>%
  mutate(across(5:ncol(.), as.numeric, .names = "num_{.col}"))
```


To calculate average market prices, we have taken the average price across all columns, assuming they represent different years, months, or days. Additionally, we compute the average price per market code (mktcode), without differentiating by crop type. This is because, for our analysis, we are interested in understanding market-level price trends rather than specific crop variations.


```{r setup7}
# Compute the average price per market code by averaging across all price columns, 
# assuming they represent different time periods (e.g., years, months, or days),
# without differentiating by crop type since it's not relevant for our analysis.
average_price <- price_data %>%
  rowwise() %>%
  mutate(avg_price = mean(c_across(5:ncol(.)), na.rm = TRUE)) %>%
  ungroup() %>%
  select(mktcode, avg_price) %>%
  group_by(mktcode) %>%
  summarise(avg_price = mean(avg_price, na.rm = TRUE))

# Print first few rows to verify the computed average prices
print(head(average_price))
```

```{r setup8}
# Merge market coordinates with the computed average prices 
# using "mktcode" as the common key to link both datasets.
merged_data <- market_sf %>%
  left_join(average_price, by = "mktcode")

# Print the first few rows to verify that the merge was successful
print(head(merged_data))
```
# Calculate minimum distances
## Coast line
```{r setup9}
# Load coastline data from the Natural Earth dataset as an sf object 
coastline <- ne_coastline(scale = "medium", returnclass = "sf")
```

```{r setup10}
# Calculate minimum distance from market places to coastline
merged_data$dist_to_coast <- apply(st_distance(merged_data, coastline), 1, min)  


# Convert distance from meters to kilometers
merged_data$dist_to_coast <- merged_data$dist_to_coast / 1000
```

### Check the minimum distance calculated
```{r setup11, warning=FALSE}
# Create graph to check if the minimum distance is correctly calculated

# Extract longitude and latitude from the geometry column of the sf object
merged_data <- merged_data %>%
  mutate(longitude = st_coordinates(.)[,1], 
         latitude = st_coordinates(.)[,2])

# Filter the dataset to only include markets within Africa's bounding box
merged_data_africa <- merged_data %>%
  filter(longitude > -25 & longitude < 60 & latitude > -40 & latitude < 40)

# Filter Coastline for Africa Only
coastline_africa <- coastline %>%
  st_crop(xmin = -25, xmax = 60, ymin = -40, ymax = 40)

# Only include Africa
africa_map <- subset(world_map, long > -25 & long < 60 & lat > -40 & lat < 40)
```

```{r setup12}
# Create map
ggplot() +
  geom_polygon(data = africa_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +  # Africa background
  geom_sf(data = coastline_africa, color = "blue", size = 0.5) +  # Coastline in blue
  geom_sf(data = merged_data_africa, aes(color = dist_to_coast), size = 2, alpha = 0.7) +  # Market points
  scale_color_viridis_c(option = "plasma") +  # Better color scale
  coord_sf(xlim = c(-25, 60), ylim = c(-40, 40), expand = FALSE) +  # Restrict to Africa
  theme_void() +  # Removes grid lines and background
  labs(title = "Market Distance to Coast", color = "Distance (km)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))  # Center title
```

### Create scatter plot that relates distance with coastline
```{r setup14}
# Create scatter plot that relates minimum distance with coast line
ggplot(merged_data, aes(x = dist_to_coast, y = avg_price)) +
  geom_point(alpha = 0.7, color = "blue") +  # Scatter points
  geom_smooth(method = "lm", color = "red", se = TRUE) +  # Linear trend line
  theme_minimal() +
  labs(title = "Market Prices vs. Minimum Distance to Coast",
       x = "Distance to Coast (km)",
       y = "Average Market Price") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The scatter plot does not reveal a clear pattern between market prices and distance to the coast. While the trend line suggests a slight positive relationship, the effect appears weak, with significant variation in prices at both short and long distances. The wide spread of data points and the broad confidence interval indicate that distance alone is not a strong predictor of market prices, suggesting that other factors, such as transportation infrastructure, local demand, or trade policies, may play a more significant role.