---
title: "Geospatial Assignment 2"
output: 
  html_document:
    toc: true          # Enable table of contents
    toc_depth: 2       # Allow headings up to level 2
    number_sections: true # Number sections and subheadings
date: '`r Sys.Date()`'
author: "Irache Vergara, Gretchen Moulton, Viktoria Gagua, Daniela Guggenberger"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
chooseCRANmirror(graphics = FALSE, ind = 1)
```

## Load libraries

```{r libraries, echo=TRUE, message=FALSE, warning=FALSE}
rm(list = ls())

# Library
library(readxl)
library(sf)
library(ggplot2)
install.packages("maps")
library(maps)
library(dplyr)
library(tidyr)
library(readr)
library(rnaturalearth)
library(rnaturalearthdata)
install.packages("osmdata")
library(osmdata)
install.packages("remotes")
library(remotes)
install.packages("afrilearndata")
#library(afrilearndata)
install.packages("RColorBrewer")
library(RColorBrewer)
library(gridExtra)
install.packages("countrycode")
library(countrycode)
```

## Load data

```{r data, echo=TRUE, message=FALSE, warning=FALSE}

# Define the base URL for raw shapefiles on GitHub
github_base <- "/vsicurl/https://raw.githubusercontent.com/Viktoriag27/Geospatial_Assignment_2/main/data"

# Define file paths for each dataset
pop <- st_read(paste0(github_base, "/ne_110m_populated_places/ne_110m_populated_places.shp"))
countries <- st_read(paste0(github_base, "/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp"))
airports <- st_read(paste0(github_base, "/ne_10m_airports/ne_10m_airports.shp"))
ports <- st_read(paste0(github_base, "/ne_10m_ports/ne_10m_ports.shp"))
locations <- st_read(paste0(github_base, "/ne_10m_populated_places_simple/ne_10m_populated_places_simple.shp"))

# Convert and transform spatial data if needed
countries <- st_as_sf(countries)
airports <- st_as_sf(airports)
airports <- st_transform(airports, crs = 4326)
ports <- st_as_sf(ports)
ports <- st_transform(ports, crs = 4326)

```
## Graphs and Visuals
To get a better feeling for the data we first plot the world map, the airports and the ports.
We decided to continue to work with airports.
```{r plots}
# Plot the world map
ggplot(data = countries) +
  geom_sf(fill = "lightblue", color = "black") +  # Fill color and borders
  theme_minimal() + 
  labs(title = "World Map")

#airports on map
ggplot() +
  # Plot world map
  geom_sf(data = countries, fill = "lightblue", color = "black") +
  # Plot airports on top
  geom_sf(data = airports, color = "yellow", size = 1) +
  theme_minimal() +
  labs(title = "World Map with Airports")

#ports on map
ggplot() +
  # Plot world map
  geom_sf(data = countries, fill = "lightblue", color = "black") +
  # Plot airports on top
  geom_sf(data = ports, color = "red", size = 1) +
  theme_minimal() +
  labs(title = "World Map with Ports")
```

## Population 

On the Natrual Earth website, we found a dataset called countries which already held the population on country level, so we first plotted that. 
```{r plot population}
ggplot(data = countries) +
  geom_sf(aes(fill = POP_EST), color = "black", size = 0.2) +  # Fill based on population
  scale_fill_viridis_b(option = "plasma", trans = "log", name = "Population", 
                       breaks = c(1e6, 1e7, 1e8, 1e9),  
                       labels = c("1M", "10M", "100M", "1B")) +  
  theme_minimal() +
  labs(title = "World Population by Country",
       subtitle = "Based on Estimated Population (pop_est)",
       caption = "Source: Natural Earth / Your Dataset") +
  theme(legend.position = "right",  
        legend.text = element_text(size = 10),  
        legend.title = element_text(size = 12, face = "bold"),  
        legend.key.height = unit(1, "cm"))  
```

Now we want to aggregate the population from the populated places/ location downloaded from the website so we get the population on country level.
In order to plot it as a map afterwards we need to get the Multipoligon objects that visualise the shape of the country as well since they are not part of the populated places data. Therefore we use the world package from R to be able to do the plots as we want them and then sum up the population numbers from all the populated places within a country.
```{r plot aggregated population}
data(world)
sf.world <- world

#Only keep the columns in sf.world df that are relevant to this analysis (iso_a2, name_long, and geom)
sf.world <- sf.world %>%
  select(iso_a2, name_long, geom, continent)

#sum up the pop by country
summed_population <- locations %>%
  group_by(iso_a2) %>%
  summarise(Total_POP_MAX = sum(pop_max, na.rm = TRUE))

summed_population_df <- summed_population %>%
  st_drop_geometry() 

world_shape_population <- sf.world %>%
  left_join(summed_population_df, by = "iso_a2") 


# Plot the heatmap
ggplot(data = world_shape_population) +
  geom_sf(aes(fill = Total_POP_MAX), color = "black", size = 0.2) +  # Fill based on population
  scale_fill_viridis_c(option = "plasma", trans = "log", na.value = "gray80",
                       name = "Total Population",  
                       breaks = c(1e6, 1e7, 1e8, 1e9),  
                       labels = c("1M", "10M", "100M", "1B")) +  
  theme_minimal() +  # Cleaner theme
  labs(
    title = "World Population by Country",
    subtitle = "Based on Aggregated Total Population"
  ) +
  theme(
    legend.position = "right",  
    legend.text = element_text(size = 10),  
    legend.title = element_text(size = 12, face = "bold"),  
    legend.key.height = unit(1, "cm"),  
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0)
  )
```


## Population Density on continent level 
Our first attempt to answer this question is to see how much of the population is made up by each country of that continent. We plot the different countries on the x-Axis and the percentege that this country makes up on that continent.
```{r hist rel}
plot_population_distribution <- function(continent_name) {
  continent_data <- countries %>%
    filter(CONTINENT == continent_name) %>%
    mutate(Pop_Percent = (POP_EST / sum(POP_EST, na.rm = TRUE)) * 100)  # Calculate % of total
  
  ggplot(continent_data, aes(x = reorder(NAME, Pop_Percent), y = Pop_Percent)) + 
    geom_bar(stat = "identity", fill = "steelblue") +  # Bar plot for % population
    labs(title = paste("Population Distribution (%) of", continent_name, "Countries"),
         x = "Country",
         y = "Percentage of Continent's Population") +
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Format as percentages
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
}

# List of continents
continents <- c("North America", "South America", "Asia", "Europe", "Africa", "Oceania")

# Apply the function to each continent and create the plots
plot_north_america <- plot_population_distribution("North America")
plot_south_america <- plot_population_distribution("South America")
plot_asia <- plot_population_distribution("Asia")
plot_europe <- plot_population_distribution("Europe")
plot_africa <- plot_population_distribution("Africa")
plot_oceania <- plot_population_distribution("Oceania")

# Display each plot separately
print(plot_north_america)
print(plot_south_america)
print(plot_asia)
print(plot_europe)
print(plot_africa)
print(plot_oceania)
```

Since we are not sure if this answer above is answering the question right, we also try another attempt, in which we plot the numbers of countries and their population.
We dont know which of these figures we are supposed to do based on the instructions.
```{r other hist}
# Perform a join based on the iso_a2 column
#summed_population <- summed_population %>%
  #left_join(sf.world[, c( "continent")], by = "iso_a2")

#delete observations with iso codes that dont make sense
summed_population <- summed_population %>%
  filter(!(iso_a2 %in% c("-99", "GS")))

# Add continent information using the countrycode library
summed_population$continent <- countrycode(summed_population$iso_a2, "iso2c", "continent")

# Re-map continent names to match the required list
summed_population <- summed_population %>%
  mutate(continent = case_when(
    continent == "Americas" & iso_a2 %in% c("US", "CA", "MX", "GT", "PA", "CU", "JM", "HN", "NI", "SV", "BZ", "CR", "DO", "HT", "BB", "TT", "BS") ~ "North America",
    continent == "Americas" ~ "South America",
    continent == "Oceania" ~ "Australia",
    TRUE ~ continent
  ))


# Define bin width 
bin_width_value <- 5000000  

# Histogram of country-level population distribution
population_plot <- ggplot(summed_population, aes(x = Total_POP_MAX, fill = continent)) +
  geom_histogram(
    binwidth = bin_width_value, 
    color = "black",
    alpha = 0.8
  ) +
  facet_wrap(~ continent, scales = "free_y") +  
  scale_fill_manual(values = rep(hist_color, 7)) +  
  scale_x_continuous(labels = label_number(scale = 1e-6, suffix = "", big.mark = ",")) + 
  x-axis
  labs(
    title = "Country Population Distribution by Continent",
    x = "Population (Millions)",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold", color = "black"),
    strip.background = element_blank(),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),  
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    legend.position = "none"  
  )

# Print the plot
print(population_plot)

```


## Locations

```{r locations}
#locations on map
ggplot() +
  # Plot world map
  geom_sf(data = countries, fill = "lightblue", color = "black") +
  # Plot airports on top
  geom_sf(data = locations, color = "yellow", size = 1) +
  theme_minimal() +
  labs(title = "World Map with Locations")
```

## Average Distances' Histogram (Airports)


```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
# Spatially join the population table with world to add country info
pop_with_country <- st_join(pop, 
                            world[, c("sovereignt", "continent")],
                            join = st_intersects)

# Remove any rows that did not join to a country
pop_with_country <- pop_with_country %>% 
  filter(!is.na(sovereignt))

# For each country, select the top 20 populated locations (using POP_MAX)
top20_pop <- pop_with_country %>%
  group_by(sovereignt) %>%                # Group by country name
  slice_max(order_by = POP_MAX,           # Order each group by the POP_MAX column
            n = 20,                     # Keep the top 20
            with_ties = FALSE) %>%       # Exclude ties beyond the 20th entry
  ungroup()

```

### Airports

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#assign each airport a country
airports <- st_transform(airports, st_crs(world))
airports_with_country <- st_join(airports, world, join = st_intersects)

#airports_with_country$sovereignt

#check how many airports per country
airport_count_by_country <- airports_with_country %>%
  group_by(sovereignt) %>%
  summarise(airport_count = n())

#print(airport_count_by_country)

airports_cleaned <- airports_with_country %>%
  filter(!is.na(sovereignt))
```
```{r}

# First, let's see what columns we have in airports_cleaned
print("Columns in airports_cleaned:")
names(airports_cleaned)

# Now let's check step by step
pop_airport_pairs <- top20_pop %>%
  group_by(sovereignt) %>%
  mutate(matching_airports = list(airports_cleaned[airports_cleaned$sovereignt == first(sovereignt), ])) %>%
  unnest(matching_airports, names_repair = "unique_quiet", names_sep = "_airport") 

# Check what columns we actually have
#print("Columns in pop_airport_pairs:")
#names(pop_airport_pairs)
```



```{r}

within_country_distances <- pop_airport_pairs %>%
  select(
    sovereignt,
    city = NAME,
    airport = matching_airports_airportname.x,
    geometry,
    geometry_airport = matching_airports_airportgeometry
  ) %>%
  rowwise() %>%  # Process each row individually
  mutate(
    distance_km = as.numeric(st_distance(geometry, geometry_airport))/1000
  ) %>%
  ungroup() %>%  # Remove rowwise grouping
  arrange(sovereignt, distance_km)
```


```{r}
country_avg_distances <- within_country_distances %>%
  group_by(sovereignt) %>%
  summarise(
    avg_distance_km = mean(distance_km)
  ) %>%
  arrange(avg_distance_km)

# Let's look at the results
head(country_avg_distances)

# Create continent-wise data
continent_distances <- country_avg_distances %>%
  left_join(
    pop_airport_pairs %>% 
      st_drop_geometry() %>%  # Remove the geometry to make it a regular dataframe
      select(sovereignt, continent) %>% 
      distinct(),
    by = "sovereignt"
  )

head(continent_distances)

```


```{r}


ggplot(continent_distances, aes(x = avg_distance_km)) +
  geom_histogram(bins = 10, fill = "red", color = "black") +
  facet_wrap(~continent) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),  
    axis.text.y = element_text(size = 8),  
    strip.text = element_text(size = 10),  # Continent labels size
    panel.grid.minor = element_blank(),    
    plot.title = element_text(size = 14)   # Title size
  ) +
  scale_x_continuous(
    breaks = seq(0, 4000, by = 1000),    
    labels = seq(0, 4000, by = 1000)      
  ) +
  scale_y_continuous(
    breaks = seq(0, 25, by = 5)           # Set specific y-axis breaks
  ) +
  labs(
    x = "Average Distance (km)",
    y = "Count",
    title = "Distribution of Average Airport Distances by Continent"
  )

## THE OUTLIERS IN THE HISTOGRAMS ARE PRESENT SINCE THERE ARE COUNTED THE TERRITORIES WHICH BELONG TO EUROPEAN COUNTRIES, FOR EXAMPLE QUEEN BEATRIZ INTERNATIONAL AIRPORT SO OUR FUNCTION CALCULATES DISTANCE BETWEEN AMSTERDAM AND DUTCH CARIBBEAN ISLAND OF ARUBA
```

## Prepare data

```{r setup2}
# Download and read the market coordinates data from a GitHub raw file
url <- "https://raw.githubusercontent.com/Viktoriag27/Geospatial_Assignment_2/main/Data/Part%202/MktCoords.xlsx"
temp_file <- tempfile(fileext = ".xlsx")  
download.file(url, temp_file, mode = "wb") 
market_data <- read_excel(temp_file)
```

```{r setup3}
# Convert the market coordinates dataframe into a spatial sf object using longitude and latitude
market_sf <- st_as_sf(market_data, coords = c("longitude", "latitude"), crs = 4326)
print(market_sf)

world_map <- map_data("world")
```


```{r setup4}
# Filter for Africa’s 
africa_map <- subset(world_map, long > -25 & long < 60 & lat > -40 & lat < 40)

# Create plot
ggplot() +
  geom_polygon(data = africa_map, aes(x = long, y = lat, group = group), fill = "gray90", color = "white") +
  geom_sf(data = market_sf, color = "red", size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Market Locations") +  # Remove x and y labels
  theme(
    axis.title = element_blank(),  # Remove axis titles
    axis.text = element_blank(),   # Remove axis text
    axis.ticks = element_blank(),   # Remove axis ticks
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
```


```{r setup5}
# Download and read the price data from a GitHub raw file
price_file <- "https://raw.githubusercontent.com/Viktoriag27/Geospatial_Assignment_2/main/Data/Part%202/PriceMaster4GAMS.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
download.file(price_file, temp_file, mode = "wb")
price_data <- read_excel(temp_file)

head(price_data)
```


```{r setup6}
# Convert price columns to numeric format, ignoring the first 4 columns 
price_data <- price_data %>%
  mutate(across(5:ncol(.), as.numeric, .names = "num_{.col}"))
```


To calculate average market prices, we have taken the average price across all columns, assuming they represent different years, months, or days. Additionally, we compute the average price per market code (mktcode), without differentiating by crop type. This is because, for our analysis, we are interested in understanding market-level price trends rather than specific crop variations.


```{r setup7}
# Compute the average price per market code by averaging across all price columns, 
# assuming they represent different time periods (e.g., years, months, or days),
# without differentiating by crop type since it's not relevant for our analysis.
average_price <- price_data %>%
  rowwise() %>%
  mutate(avg_price = mean(c_across(5:ncol(.)), na.rm = TRUE)) %>%
  ungroup() %>%
  select(mktcode, avg_price) %>%
  group_by(mktcode) %>%
  summarise(avg_price = mean(avg_price, na.rm = TRUE))

# Print first few rows to verify the computed average prices
print(head(average_price))
```

```{r setup8}
# Merge market coordinates with the computed average prices 
# using "mktcode" as the common key to link both datasets.
merged_data <- market_sf %>%
  left_join(average_price, by = "mktcode")

# Print the first few rows to verify that the merge was successful
print(head(merged_data))
```

## Calculate Minimum Distances to Market Places

### Coast line

```{r setup9}
# Load coastline data from the Natural Earth dataset as an sf object 
coastline <- ne_coastline(scale = "medium", returnclass = "sf")
```

```{r setup10}
# Calculate minimum distance from market places to coastline
merged_data$dist_to_coast <- apply(st_distance(merged_data, coastline), 1, min)  


# Convert distance from meters to kilometers
merged_data$dist_to_coast <- merged_data$dist_to_coast / 1000
```

### Check the minimum distance calculated

```{r setup11, warning=FALSE}
# Create graph to check if the minimum distance is correctly calculated

# Extract longitude and latitude from the geometry column of the sf object
merged_data <- merged_data %>%
  mutate(longitude = st_coordinates(.)[,1], 
         latitude = st_coordinates(.)[,2])

# Filter the dataset to only include markets within Africa's bounding box
merged_data_africa <- merged_data %>%
  filter(longitude > -25 & longitude < 60 & latitude > -40 & latitude < 40)

# Filter Coastline for Africa Only
coastline_africa <- coastline %>%
  st_crop(xmin = -25, xmax = 60, ymin = -40, ymax = 40)

# Only include Africa
africa_map <- subset(world_map, long > -25 & long < 60 & lat > -40 & lat < 40)
```

```{r setup12}
# Create map
ggplot() +
  geom_polygon(data = africa_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +  # Africa background
  geom_sf(data = coastline_africa, color = "blue", size = 0.5) +  # Coastline in blue
  geom_sf(data = merged_data_africa, aes(color = dist_to_coast), size = 2, alpha = 0.7) +  # Market points
  scale_color_viridis_c(option = "plasma") +  # Better color scale
  coord_sf(xlim = c(-25, 60), ylim = c(-40, 40), expand = FALSE) +  # Restrict to Africa
  theme_void() +  # Removes grid lines and background
  labs(title = "Market Distance to Coast", color = "Distance (km)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))  # Center title
```

### Create scatter plot that relates average prices with coastline

```{r setup14}
# Create scatter plot that relates average prices with distance to coast line
ggplot(merged_data, aes(x = dist_to_coast, y = avg_price)) +
  geom_point(alpha = 0.7, color = "blue") +  # Scatter points
  geom_smooth(method = "lm", color = "red", se = TRUE) +  # Linear trend line
  theme_minimal() +
  labs(title = "Market Prices vs. Minimum Distance to Coast",
       x = "Distance to Coast (km)",
       y = "Average Market Price") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The scatter plot does not reveal a clear pattern between market prices and distance to the coast. While the trend line suggests a slight positive relationship, the effect appears weak, with significant variation in prices at both short and long distances. The wide spread of data points and the broad confidence interval indicate that distance alone is not a strong predictor of market prices, suggesting that other factors, such as transportation infrastructure, local demand, or trade policies, may play a more significant role.

### Roads
### Calculate the minimum distance from markets to nearest roads
We now gather road data from Natural Earth. We filter on roads in Africa using the continent file and calculate the nearest road by first calculating the distance from each market to each road. Then we store the nearest roads with the shortest distances. We extract these roads and calculate the distance from market to nearest road in km. Finally we inspect our data to make sure the process ran smoothly. 
```{r}
# Define the URL to the shapefile on GitHub
shp_url <- "/vsicurl/https://raw.githubusercontent.com/Viktoriag27/Geospatial_Assignment_2/main/Data/ne_10m_roads/ne_10m_roads.shp"
# Read the shapefile directly from the URL
road_data <- st_read(shp_url)

#Filter on raods in Africa
road_data<- road_data %>%
  filter(continent == "Africa")

# Calculate the distance between each market point and all road geometries
distances <- st_distance(merged_data, road_data)

# Identify the nearest road for each market
nearest_road_index <- apply(distances, 1, which.min)

# Extract the nearest road geometries
nearest_roads <- road_data[nearest_road_index, ]

# Add the nearest road geometry to the merged_data dataframe
merged_data$nearest_road_geom <- nearest_roads$geometry

# Calculate the distance to the nearest road in km
merged_data$distance_to_nearest_road <- as.numeric(apply(distances, 1, min)/1000)

# View our changes in merged_data
head(merged_data)
```

For fun, we map the distance from markets to the nearest road. 

```{r}
# Create map
ggplot() +
  geom_polygon(data = africa_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +  # Africa background
  geom_sf(data = road_data, color = "green", size = 0.5) + 
  geom_sf(data = merged_data, aes(color = distance_to_nearest_road), size = 2, alpha = 0.7) +  # Market points
  scale_color_viridis_c(option = "cividis") +  # Better color scale
  coord_sf(xlim = c(-25, 60), ylim = c(-40, 40), expand = FALSE) +  # Restrict to Africa
  theme_void() +  # Removes grid lines and background
  labs(title = "Market Distance to the Nearest Road", color = "Distance (km)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))  # Center title
```

### Create scatter plot that relates average prices with nearest road

```{r}
# Create scatter plot that relates average prices with distance to nearest road
ggplot(merged_data, aes(x = distance_to_nearest_road, y = avg_price)) +
  geom_point(alpha = 0.7, color = "blue") +  # Scatter points
  geom_smooth(method = "lm", color = "green", se = TRUE) +  # Linear trend line
  theme_minimal() +
  labs(title = "Market Prices vs. Minimum Distance to Road",
       x = "Distance to the Nearest Road (km)",
       y = "Average Market Price") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

From this scatter plot, we detect no relationship between the average price of goods at the market and the distance to the nearest road. The trend line appears to have a positive slope but this is due to an outlier. Most markets are relatively close to roads and thus there are no identifiable trends in market pricing due to road access. 

### Airports

To gather data on airports in Africa, I used open source data from the afrimapr project and filtered on medium and large airports in Africa. I originally tried to use natural earth data; however, as there was no continent variable, I attempted to filter on airports in Africa by filtering on the bounding box. This created issues when I mapped my data as it also showed airports outside of Africa, such as on the Arabian Peninsula, that were within the bounding box. While it did not impact the distance calculation, it aesthetically did not look as organized or clear. I also tried to use st_within or st_intersect but encountered issues. Using the afrimapr data allowed for the simplest method and allowed me to map the markets and airports, which adds additional understanding to this project. 

```{r}
# Load the dataset for African airports
data("afriairports")

# We select relevant columns, filter on large and medium airports, ensure the geometry is treated as points, and that the correct CRS is being used
africa_airports <- afriairports %>%
  select(name, country_name, geometry,type) %>%
  filter(type %in% c("large_airport", "medium_airport")) %>% 
  st_as_sf() %>%  # Ensure it remains an sf object
  st_transform(crs = 4326)  # Optional: Set CRS to WGS84 (if needed)

# View the first few rows to ensure its correct
print(head(africa_airports))
```

### Calculate the minimum distance from markets to nearest airport

```{r}
# Calculate minimum distance from market places to airport
merged_data$dist_to_airport <- apply(st_distance(merged_data, africa_airports), 1, min)  

# Convert distance from meters to kilometers
merged_data$dist_to_airport <- merged_data$dist_to_airport / 1000

# View the results of our calculation
print(head(merged_data))
```
We map the distance from markets to the nearest airport.  
```{r}
library(ggplot2)
# Create map
ggplot() +
  geom_polygon(data = africa_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +  # Africa background
  geom_sf(data = africa_airports, color = "red", size = 0.5) +  # Airports in blue
  geom_sf(data = merged_data, aes(color = dist_to_airport), size = 2, alpha = 0.7) +  # Market points
  scale_color_viridis_c(option = "magma") +  # Better color scale
  coord_sf(xlim = c(-25, 60), ylim = c(-40, 40), expand = FALSE) +  # Restrict to Africa
  theme_void() +  # Removes grid lines and background
  labs(title = "Market Distance to Airport", color = "Distance (km)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))  # Center title
```

### Create scatter plot that relates average prices with nearest airport

```{r}
# Create scatter plot that relates average prices with distance to airports
ggplot(merged_data, aes(x = dist_to_airport, y = avg_price)) +
  geom_point(alpha = 0.7, color = "blue") +  # Scatter points
  geom_smooth(method = "lm", color = "pink", se = TRUE) +  # Linear trend line
  theme_minimal() +
  labs(title = "Market Prices vs. Minimum Distance to Airport",
       x = "Distance to Airport (km)",
       y = "Average Market Price") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Observing our results from the scatter plot, it is unclear if there is a relationship between the average price of goods at the market and the distance to the nearest airport. The trend line moves slightly downward, indicating a slight negative correlation as with price as distance increases. However, our data contains many higher and lower contradictory outliers. 