---
title: "Geospatial Data Science Assignment 3 - Task 1"
author: "Daniela and Viktoria"
date: "2025-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1

## Load libraries

```{r}
# install.packages(c("tidyverse", "sf", "raster", "SPEI", "ncdf4", "tigris", "ggplot2", "sf", "dplyr", "ncdf4"))
library(tidyverse)   # Data manipulation
library(sf)          # Spatial data handling
library(raster)      # Working with raster climate data
library(SPEI)        # Calculating SPEI
library(ncdf4)       # Reading NetCDF climate data
library(tigris)
library(ggplot2)
library(sf)
library(dplyr)
library(ncdf4)
```

## Load Data and Plot Mainland US
We use data about the information of the US-states from the tigris library. Before doing anything else, we restrict the states we use, so it matches the instructions (exclude Alasska, Hawaii, smaller islands, ...).
```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
Sys.setenv("VROOM_SHOW_PROGRESS" = "false") # Disables progress bar

# Load US states
us_states <- suppressMessages(suppressWarnings(states(cb = TRUE, year = 2023)))

#Since the US also has some islands and Alaska as part of it which are NOT part 
#of the Regions we are supposed to create, we have to get rid of them 
us_contiguous <- us_states %>%
  filter(!STUSPS %in% c("AK", "HI", "PR", "GU", "VI", "MP", "AS"))
```

```{r}
# Plot the cleaned-up contiguous U.S.
ggplot(data = us_contiguous) +
  geom_sf(fill = "lightblue", color = "black") +
  labs(title = "Mainland US") +
  theme_minimal()
```

## Data Preprocessing

Since we are asked to perform regional level operations, we split the states into northeast, midwest, south and west. 

```{r}
#Since the instructions ask us to recieve a SPEI index on REGIONAL level, we 
#have to split up our states in the wanted regions
us_regions <- us_contiguous %>%
  mutate(region = case_when(
    STUSPS %in% c("ME", "NH", "VT", "MA", "RI", "CT", "NY", "NJ", "PA") ~ "Northeast",
    STUSPS %in% c("OH", "IN", "IL", "MI", "WI", "MN", "IA", "MO", "ND", "SD", "NE", "KS") ~ "Midwest",
    STUSPS %in% c("DE", "MD", "DC", "VA", "WV", "KY", "NC", "SC", "GA", "FL", 
                  "TN", "AL", "MS", "AR", "LA", "OK", "TX") ~ "South",
    STUSPS %in% c("MT", "ID", "WY", "CO", "NM", "AZ", "UT", "NV", "WA", "OR", "CA") ~ "West",
    TRUE ~ NA_character_ # Ensure no other states are included
  ))

# Check the data to confirm regions
table(us_regions$region)

#plot US divided in the regions
ggplot(data = us_regions) +
  geom_sf(aes(fill = region), color = "black") +
  scale_fill_manual(values = c(
    "Northeast" = "#D55E00",  
    "Midwest" = "#009E73",   
    "South" = "#56B4E9",     
    "West" = "#CC79A7"
  )) +
  labs(title = "U.S. Regions", fill = "Region") +
  theme_minimal()
```

## US Regions as Combined Polygons

To perform calculations on regional level, we combine the states within a region together to one polygon each. 

```{r}
#Now we have the regions we actually have to combine the states within a region 
#together so we end up with only 1 polygon per region
us_regions_combined <- us_regions %>%
  group_by(region) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

# Check the result
#print(us_regions_combined)

#plot combined regions
ggplot(data = us_regions_combined) +
  geom_sf(aes(fill = region), color = "black") +
  scale_fill_manual(values = c(
    "Northeast" = "#D55E00",  
    "Midwest" = "#009E73",   
    "South" = "#56B4E9",     
    "West" = "#CC79A7"
  )) +
  labs(title = "U.S. Regions as Combined Polygons", fill = "Region") +
  theme_minimal()

```

## Filtering 50 years and Cropping SPEI Data for US Regions

For the Speix data we need to extract the last 50 years of observations and allingn them to the region they belong to. 
```{r}
#Load the SPEI index data 
spei_data <- brick("~/Desktop/Victoria/DSDM/Term2/Geospatial/Assignment_3/PS3/data/spei01.nc")

# Check the dimensions of the raster brick 
print(dim(spei_data))

# Get the time values (this should correspond to each layer in spei_data)
time_values <- getZ(spei_data)

# Extract the date part of the names (remove 'X' and convert to Date)
time_dates <- as.Date(gsub("^X", "", names(spei_data)), format="%Y.%m.%d")

# Check the result
#print(time_dates)


#The instructions wants us to only look at the last 50 years
#To filter out the last 50 years we need to check 
#tail(time_values) # We see the data is monthly and the last entry was "2023-12-16"
#Therefore to get the last 50 years (600months) we need to do the following
#set an end date
end_date <- as.Date("2023-12-16")
#and a starting date 
start_date <- end_date - months(600)

# Filter the data to the dates we just found
time_filtered <- time_dates[time_dates >= start_date & time_dates <= end_date]

# Find the layer indices for the filtered time values
matching_layers <- which(time_dates %in% time_filtered)

# Subset the raster data using the matching layers
spei_filtered <- spei_data[[matching_layers]]

# Check the dimensions of the filtered raster data
print(dim(spei_filtered))


# Now we have only the last 50 years of SPEIX data left, but now we still need to crop it 
# so we only have to data that matches the US map we already have left.
# seperating in Regions

# Ensure your us_contiguous object is in the correct coordinate system (CRS) for the raster data
us_contiguous <- st_transform(us_contiguous, crs = crs(spei_filtered))

# Crop the spei_filtered raster to the extent of the US states (us_contiguous)
spei_cropped <- crop(spei_filtered, extent(us_contiguous))

plot(spei_cropped)

# Crop the spei_filtered raster to the extent of the combined regions
spei_cropped_regions <- crop(spei_filtered, extent(us_regions_combined))

# Plot the cropped raster data for the regions
#plot(spei_cropped_regions)
print(dim(spei_cropped_regions))


# If using regions, you can do the same:
spei_masked_regions <- mask(spei_cropped_regions, us_regions_combined)

# Plot the masked raster data
plot(spei_masked_regions)

print(dim(spei_masked_regions))

#SPEIX by region 
# Mask the cropped raster by the region
spei_masked_northeast <- mask(spei_cropped, us_regions_combined[us_regions_combined$region == "Northeast", ])
spei_masked_midwest <- mask(spei_cropped, us_regions_combined[us_regions_combined$region == "Midwest", ])
spei_masked_south <- mask(spei_cropped, us_regions_combined[us_regions_combined$region == "South", ])
spei_masked_west <- mask(spei_cropped, us_regions_combined[us_regions_combined$region == "West", ])

#just to check
#print(dim(spei_masked_midwest))
#plot(spei_masked_northeast)
```

## Panel Data

```{r}
#-------------------------------
# Calculate the mean SPEI index for each region at each time step
#-------------------------------

# Create a time vector corresponding to the filtered dates
time_vector <- time_filtered

# Calculate the mean SPEI (ignoring NA values) for each time layer and region
spei_ne_mean <- cellStats(spei_masked_northeast, stat = 'mean', na.rm = TRUE)
spei_mw_mean <- cellStats(spei_masked_midwest, stat = 'mean', na.rm = TRUE)
spei_s_mean  <- cellStats(spei_masked_south, stat = 'mean', na.rm = TRUE)
spei_w_mean  <- cellStats(spei_masked_west, stat = 'mean', na.rm = TRUE)

#---------------------------------------------
#   Build a "panel" data frame
#    date x region -> SPEI time series
#---------------------------------------------
spei_df <- data.frame(
  date   = rep(time_vector, times = 4),
  SPEI   = c(spei_ne_mean, spei_mw_mean, spei_s_mean, spei_w_mean),
  region = rep(c("Northeast", "Midwest", "South", "West"), each = length(time_vector))
)

# Calculate the overall average SPEI across all regions for each time step
overall_df <- spei_df %>%
  group_by(date) %>%
  summarize(overall_SPEI = mean(SPEI, na.rm = TRUE))

#overall_df
```

```{r}
# Quick check of the panel structure
head(spei_df)
tail(spei_df)

```

## Evolution of SPEI Index

Plotting the evolution of SPEI for each region and the overall trend

```{r}

## Final plot
p <- ggplot() +
  # Plot individual region time series
  geom_line(data = spei_df, aes(x = date, y = SPEI, color = region)) +
  # Add a smoothed line for the overall average across regions
  geom_smooth(
    data   = overall_df, 
    aes(x = date, y = overall_SPEI),
    color  = "black", 
    linewidth = 0.8, 
    se     = FALSE, 
    method = "loess",
    span   = 0.4    # <-- smaller span reduces the smoothing
  ) +
  scale_y_continuous(
    breaks = c(-1, 0, 1)  # Only show ticks at -1, 0, and 1
  ) + 
  labs(
    title    = "Evolution of SPEI Index by Region",
    subtitle = "Black line shows the overall smoothed average across regions",
    x        = "Date", 
    y        = "SPEI Index",
    color    = "Region"
  ) +
  theme_minimal()

print(p)
```
